Discrete Quantum Theories and Computing

### 2018-11-22 dissertation.tex
dissertation.lyx "2.2 Quantum Probability" Updating...

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[chapter]
\newtheorem{thm}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{case}{Case}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{question}{Question}[chapter]
\newtheorem*{method}{Method}
\newtheorem{example}{Example}[chapter]
\newtheorem{postulate}{Postulate}[chapter]

@@

We first replace the continuum of complex numbers by discrete finite
fields in the quantum theory. The simplest theory, defined over unrestricted
finite fields, is so weak that it cannot express Deutsch’s algorithm.
This quantum theory is, however, also so powerful that it can be used
to solve the $\usat$ problem, which is as hard as a general $\NP$-complete
problem. The second framework employs only finite fields with no solution
to $x^{2}+1=0$, and thus permits an elegant complex representation
of the extended field by adjoining $\rmi=\sqrt{-1}$. Because the
states of a discrete $n$-qubit system are in principle enumerable,
we can count the number of states, and determine the proportions of
entangled and unentangled states. Although this improved framework
allows the probabilistic Grover search algorithm in a local region,
we haven't found sensible correspondences between quantum probability
measures and the whole state space in general.

@@

Since the previous abstraction doesn't specify the process to generate
the outcomes, this process could well be a quantum experiment. Let
us prepare a beam of one kind of spin~$1$ particles with basis vectors
$\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle, this beam can
be split by a Stern-Gerlach type experiment according to the eigenvalues
of the observable $\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}$,
and the states corresponding to the split beams are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of sending a beam, if only one particle is sent to the beam
splitter, the state after the experiment is one of $\ket{0}$, $\ket{1}$,
and $\ket{2}$, and there is a probability for each post-experimental
state. In the language of our abstraction, the set of outcomes is
$\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $. All possible
events are $\emptyset$, $\left\{ \ket{0}\right\} $, $\left\{ \ket{1}\right\} $,
$\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. And this experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
similar to the previous one. While sending exactly the same particle,
we use a different beam splitter corresponding to the observable $\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$,
where $\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ and $\ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}$.
Although the sample space $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $
and the probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$
defined by this experiment is different from the previous one, these
two experiments may produce the same post-experimental state~$\ket{2}$,
and the probability of the common event $\left\{ \ket{2}\right\} $
is believed to be the same, i.e., $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
The general situation can be summarized as the following postulate.

@@

\subsection{The Born Rule and Gleason's Theorem\label{subsec:Gleason's-Theorem-and}}

After introducing quantum probability measures, we might follow the
convention to introduce the Born rule which is the only way to construct
quantum probability measures in CQT. However, since the variants of
quantum probability measures in Sec.~\ref{sec:Toward-IVPM} and Chapter~\ref{chap:QIVPM}
could not be constructed by the Born rule easily, it is instructive
to search for a Born rule here, and we will actually find the unique
one.

Although quantum probability measures constructed by gluing classical
probability measures in the previous section looks complex, it could
actually be constructed by operators according to Gleason's theorem.

@@

Although Gleason's theorem can correspond every quantum probability
measure to a unique mixed state~$\rho$, it doesn't explain whether
this state $\rho$ characterizing the state of particle sending to
a Stern-Gerlach type experiment. For example, a variant of quantum
probability measure has only one instance defined in Eq.~\ref{eq:QPMFF37}.
This instance predicts all outcomes to be equally probable which intuitively
corresponds to the mixed state $\frac{1}{3}\proj{0}+\frac{1}{3}\proj{1}+\frac{1}{3}\proj{2}$
in its space. However, ignoring the input state and always predicting
equally probable is not a sensible quantum theory.

Let $\ket{\Phi}$ characterize the state of particle sending to a
Stern-Gerlach type experiment, and $\muB_{\Phi}$ be the corresponding
quantum probability measure of the resulting events.

To find a sensible correspondences between quantum probability measures
and the input states,

\noindent , when sending a particle to a Stern-Gerlach type experiment,
we will get a quantum probability measure of the form $\Tr\left(\rho P\right)$
for some operator~$\rho$.

Recall in Sec.~\ref{CQC1qubitBloch.sec}, the state of the particle
can be characterized by a pure state $\ket{\Psi}\in\Hilb$ or more
generally a mixed state~$\rho'$. Although the function $\Tr\left(\rho P\right)$
is parametrized by

\todo{Rewrite the beginning of this paragraph to explain the meaning
of the Born rule... Maybe move Gleason's Theorem here?} A conventional
quantum probability measure can easily be constructed using the Born
rule if one knows the current pure unnormalized quantum state $\ket{\Phi}\in\mathcal{H}$;
then the Born rule induces a probability measure $\muB_{\Phi}$ defined
as 
\begin{equation}
\muB_{\Phi}(P)=\frac{\melem{\Phi}{P}{\Phi}}{\ip{\Phi}{\Phi}}\,.\label{eq:Born}
\end{equation}
If $\ket{\Phi}$ is normalized, Eq.~(\ref{eq:Born}) could be simplified
as $\muB_{\Phi}(P)=\melem{\Phi}{P}{\Phi}$. For mixed states $\rho=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}$,
where $\ket{\Phi_{j}}\in\mathcal{H}$ are normalized, $q_{j}>0$,
and $\sum_{j=1}^{N}q_{j}=1$, the generalized Born rule induces a
probability measure $\muB_{\rho}$ defined as \cite{Born1983,Mermin2007,Jaeger2007}
\begin{equation}
\muB_{\rho}\left(P\right)=\Tr\left(\rho P\right)=\sum_{j=1}^{N}q_{j}\muB_{\Phi_{j}}\left(P\right)\,.\label{eq:mixed-Born}
\end{equation}

As an example, consider a three-dimensional Hilbert space with orthonormal
basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $ and an observable
$\mathbf{O}$ with spectral decomposition $\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
i.e., the $i$th eigenvalue $\lambda_{i}=i$ and the $i$th eigenprojector
$P_{i}=\proj{i-1}$. Two fragments of valid probability measures~$\mu_{1}$
and $\mu_{2}$ that can be associated with this space are defined
in Table~\ref{tab:quantum-probability-measure}. By the Born rule,
the first probability measure corresponds to the quantum system being
in the pure state $\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ and the second
corresponds to the quantum system being in the state $\frac{\proj{0}+\proj{2}}{2}$.
The expectation values of the observable $\mathbf{O}$, $\expval{\mathbf{O}}_{\mu_{1,2}}$,
are $1.5$ in the first case and $2$ in the second.
\begin{table}
\noindent \begin{centering}
\caption{\label{tab:quantum-probability-measure}Two fragments of valid probability
measures~$\mu_{1}$ and $\mu_{2}$.}
\par\end{centering}
\noindent \centering{}%
\begin{tabular}{ccccccccccc}
\toprule 
\addlinespace
$\ket{\Psi}$ & $\ket{0}$ & $\ket{1}$ & $\ket{2}$ & $\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ & $\frac{\ket{0}+\rmi\ket{1}}{\sqrt{2}}$ & $\frac{\ket{0}+\ket{2}}{\sqrt{2}}$ & $\frac{\ket{0}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{\ket{1}+\ket{2}}{\sqrt{2}}$ & $\frac{\ket{1}+\rmi\ket{2}}{\sqrt{2}}$ & \ldots\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
$\mu_{1}\left(\proj{\Psi}\right)$ & $\frac{1}{2}$ & $\frac{1}{2}$ & $0$ & $1$  & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & \ldots\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\mu_{2}\left(\proj{\Psi}\right)$ & $\frac{1}{2}$ & $0$ & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{2}$ & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & \ldots\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}
\todo{Make the meaning of Gleason's Theorem more clear? So that we
can explain the motivation of the question in Chapter~\ref{chap:Further-Discussion}.}
A conventional quantum probability measure can be easily constructed
from a state $\rho$ according to the Born rule; conversely, this
is the only way to obtain a conventional quantum probability measure
according to Gleason's theorem~\cite{gleason1957,Redhead1987-REDINA,peres1995quantum}.

Although the Born rule is widely be considered a postulate of quantum
theory, it could be derived from a smaller set of abstract definitions
and axioms according to Gleason's theorem~\cite{peres1995quantum}.
Since the Born rule linking the probability amplitudes and probability
might not hold naively when probability amplitudes is in finite fields
or the probability becomes an interval. Considering a Gleason-like
theorem is an appealing method to search a rule computing probability
in all these cases.

Although Gleason's theorem associates every quantum probability measure~$\mu$
to a mixed state~$\rho$, it is not obvious how to find $\rho$ for
a given $\mu$. While hard in general, this task can be done easily
if $\mu$ satisfies $\mu\left(P\right)=1$ for some projection~$P$
when the dimension $D\geq3$. According to Gleason's theorem, there
exists a unique mixed state $\rho=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}$
such that 
\begin{equation}
1=\mu\left(P\right)=\muB_{\rho}\left(P\right)=\sum_{j=1}^{N}q_{j}\muB_{\Phi_{j}}\left(P\right)=\sum_{j=1}^{N}q_{j}\melem{\Phi_{j}}{P}{\Phi_{j}}\,.
\end{equation}
Together with the fact that $0\le\melem{\Phi_{j}}{P}{\Phi_{j}}\le1$,
$q_{j}>0$, and $\sum_{j=1}^{N}q_{j}=1$, we must conclude that $\melem{\Phi_{j}}{P}{\Phi_{j}}=1$
for all~$j$, and thus $P\ket{\Phi_{j}}=\ket{\Phi_{j}}$. Hence,
\begin{equation}
P\rho=\sum_{j=1}^{N}q_{j}P\proj{\Phi_{j}}=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}=\rho\,.
\end{equation}
Conversely, $P\rho=\rho$ also implies $\muB_{\rho}\left(P\right)=1$
because the trace of any mixed state is $1$. The whole idea can then
be summarized as the following proposition.

\begin{prop}\label{prop:Born-probability-one}$\muB_{\rho}\left(P\right)=1$
if and only if $P\rho=\rho$.\end{prop}

The final property is that the Born rule is basis independent~\cite{peres1995quantum,544199}:
\todo{Update the basis independent part with more details...}

\begin{prop}\label{prop:Born-unitary-invariant}$\muB_{U\rho U^{\dagger}}\left(UPU^{\dagger}\right)=\muB_{\rho}\left(P\right)$,
where $U$ is any unitary map.\end{prop}

In Sec.~\ref{sec:Toward-IVPM}, we will use an analog of Gleason's
theorem and these two properties to deduce that there is no “sensible”
real-valued Born rule over finite fields.

>>>>

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[chapter]
\newtheorem{thm}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{case}{Case}
\newtheorem{corollary}{Corollary}[chapter]

\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{question}{Question}[chapter]
\newtheorem*{method}{Method}
\newtheorem{example}{Example}[chapter]
\newtheorem{postulate}{Postulate}[chapter]

@@

We first replace the continuum of complex numbers by discrete finite
fields in the quantum theory. The simplest theory, defined over unrestricted
finite fields, is so weak that it cannot express Deutsch's algorithm.
This quantum theory is, however, also so powerful that it can be used
to solve the $\usat$ problem, which is as hard as a general $\NP$-complete
problem. The second framework employs only finite fields with no solution
to $x^{2}+1=0$, and thus permits an elegant complex representation
of the extended field by adjoining $\rmi=\sqrt{-1}$. Because the
states of a discrete $n$-qubit system are in principle enumerable,
we can count the number of states, and determine the proportions of
entangled and unentangled states. Although this improved framework
allows the probabilistic Grover search algorithm in a local region,
we haven't found sensible correspondences between quantum probability
measures and the whole state space in general.

@@

Since the previous abstraction doesn't specify the process to generate
the outcomes, this process could well be a quantum experiment. Let
us prepare a beam of one kind of spin~$1$ particles with basis vectors
$\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle, this beam can
be split by a Stern-Gerlach type experiment according to the eigenvalues
of the observable $\mathbf{O}_{0}$ with with spectral decomposition
\begin{equation}
\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}\,,\label{eq:observable}
\end{equation}
and the states corresponding to the split beams are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of sending a beam, if only one particle is sent to the beam
splitter, the state after the experiment is one of $\ket{0}$, $\ket{1}$,
and $\ket{2}$, and there is a probability for each post-experimental
state. In the language of our abstraction, the set of outcomes is
$\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $. All possible
events are $\emptyset$, $\left\{ \ket{0}\right\} $, $\left\{ \ket{1}\right\} $,
$\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. And this experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
similar to the previous one. While sending exactly the same particle,
we use a different beam splitter corresponding to the observable $\mathbf{O}_{1}$
with spectral decomposition $\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$,
where 
\begin{align}
\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}\,, &  & \ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}\,.\label{eq:psms}
\end{align}
Although the sample space $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $
and the probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$
defined by this experiment is different from the previous one, these
two experiments may produce the same post-experimental state~$\ket{2}$,
and the probability of the common event $\left\{ \ket{2}\right\} $
is believed to be the same, i.e., $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
The general situation can be summarized as the following postulate.

@@

\subsection{Gleason's Theorem and The Born Rule\label{subsec:Gleason's-Theorem-and}}

After introducing quantum probability measures, we might follow the
convention to introduce the Born rule which is the only way corresponding
a state to construct quantum probability measures in CQT. However,
since the variants of quantum probability measures in Sec.~\ref{sec:Toward-IVPM}
and Chapter~\ref{chap:QIVPM} could not be constructed by the Born
rule easily, it is instructive to search for a Born rule here, and
we will actually find the unique one.

Although quantum probability measures constructed by gluing together
the classical ones looks complex, they could all be induced by operators
according to Gleason's theorem.

@@

Although Gleason's theorem can correspond every quantum probability
measure to a unique mixed state~$\rho$, it doesn't explain whether
this state $\rho$ characterizing the state of particle sending to
a Stern-Gerlach type experiment. For example, in contrast with CQT,
if a quantum theory always ignore the input state and predict the
experimental outcomes are equally probable, even if this quantum theory
maps every input state to a quantum probability measure, it is still
not very sensible.

Let a pure unnormalized state $\ket{\Phi}\in\mathcal{H}$ characterize
the particle sending to a Stern-Gerlach type experiment, and $\muB_{\Phi}$
be the corresponding quantum probability measure of the resulting
events. To define a sensible correspondence, we hope that if the state~$\ket{\Phi}$
is one of the outcome of a quantum event~$P$, i.e., $P\ket{\Phi}=\ket{\Phi}$,
then the event~$P$ always happens, $\muB_{\Phi}\left(P\right)=1$
and vice versa. Moreover, since the physical phenomena exists and
should be the same no matter how we describe them, the probability
of an event should be invariant despite how we choose the basis. Because
changing to another basis is the same as applying an unitary map~$U$,
we should have$\muF_{U\ket{\Phi}}\left(UPU^{\dagger}\right)=\muF_{\Phi}\left(P\right)$.
It is easy to check the correspondence satisfying the conditions in
this paragraph is unique, 
\begin{equation}
\muB_{\Phi}(P)=\frac{\melem{\Phi}{P}{\Phi}}{\ip{\Phi}{\Phi}}\label{eq:Born}
\end{equation}
and called the Born rule \cite{Born1983,Mermin2007,Jaeger2007}. If
$\ket{\Phi}$ is normalized, Eq.~(\ref{eq:Born}) could be simplified
as $\muB_{\Phi}(P)=\melem{\Phi}{P}{\Phi}$. For mixed states $\rho=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}$,
where $\ket{\Phi_{j}}\in\mathcal{H}$ are normalized, $q_{j}>0$,
and $\sum_{j=1}^{N}q_{j}=1$, the generalized Born rule induces a
probability measure $\muB_{\rho}$ defined as 
\begin{equation}
\muB_{\rho}\left(P\right)=\Tr\left(\rho P\right)=\sum_{j=1}^{N}q_{j}\muB_{\Phi_{j}}\left(P\right)\,.\label{eq:mixed-Born}
\end{equation}

As an example, consider a three-dimensional Hilbert space with orthonormal
basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $ and an observable
$\mathbf{O}_{0}$ defined in Eq.~(\ref{eq:observable}). Two fragments
of valid probability measures~$\mu_{1}$ and $\mu_{2}$ that can
be associated with this space are defined in Table~\ref{tab:quantum-probability-measure}.
By the Born rule, the first probability measure corresponds to the
quantum system being in the pure state $\ket{\ps}$ defined in Eq.~(\ref{eq:psms})
and the second corresponds to the quantum system being in the state
$\frac{\proj{0}+\proj{2}}{2}$. The expectation values of the observable
$\mathbf{O}$, $\expval{\mathbf{O}}_{\mu_{1,2}}$, are $1.5$ in the
first case and $2$ in the second.
\begin{table}
\noindent \begin{centering}
\caption{\label{tab:quantum-probability-measure}Two fragments of valid probability
measures~$\mu_{1}$ and $\mu_{2}$.}
\par\end{centering}
\noindent \centering{}%
\begin{tabular}{ccccccccccc}
\toprule 
\addlinespace
$\ket{\Psi}$ & $\ket{0}$ & $\ket{1}$ & $\ket{2}$ & $\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ & $\frac{\ket{0}+\rmi\ket{1}}{\sqrt{2}}$ & $\frac{\ket{0}+\ket{2}}{\sqrt{2}}$ & $\frac{\ket{0}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{\ket{1}+\ket{2}}{\sqrt{2}}$ & $\frac{\ket{1}+\rmi\ket{2}}{\sqrt{2}}$ & \ldots\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
$\mu_{1}\left(\proj{\Psi}\right)$ & $\frac{1}{2}$ & $\frac{1}{2}$ & $0$ & $1$  & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & \ldots\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\mu_{2}\left(\proj{\Psi}\right)$ & $\frac{1}{2}$ & $0$ & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & $\frac{1}{2}$ & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & \ldots\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}
